```{r}
#| label: setup
#| include: false

source(here::here("R/quarto-setup.R"))
```

<!-- badges: start -->
[![Project Status: Inactive – The project has reached a stable, usable state but is no longer being actively developed; support/maintenance will be provided as time allows.](https://www.repostatus.org/badges/latest/inactive.svg)](https://www.repostatus.org/#inactive)
[![License: MIT](https://img.shields.io/badge/license-MIT-green)](https://choosealicense.com/licenses/mit/)
<!-- badges: end -->

## Overview

This document focuses on demonstrating the [SIR model](https://en.wikipedia.org/wiki/Compartmental_models_in_epidemiology), originally introduced by Kermack and McKendrick in [-@kermack1927]. The SIR model is a foundational framework in epidemiology, designed to analyze the spread of infectious diseases by categorizing the population into three compartments: Susceptible (S), Infected (I), and Recovered (R).

The dynamics of the model are represented by the following set of first-order, nonlinear differential equations:

$$
\begin{aligned}
\frac{dS}{dt} &= -\beta S I \\
\frac{dI}{dt} &= \beta S I - \gamma I \\
\frac{dR}{dt} &= \gamma I
\end{aligned}
$$

where, 

- $\beta$ represents the transmission rate;
- $\gamma$ represents the recovery rate.

## Setting up the environment

```{r}
#| eval: false
#| output: false

library(checkmate, quietly = TRUE)
library(deSolve, quietly = TRUE)
library(dplyr, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(latex2exp, quietly = TRUE)
library(magrittr, quietly = TRUE)
```

## Numerical solution of the equations

```{r}
sir <- function(
    s = 1, 
    i = 0.05, 
    r = 0, 
    beta  = 3, 
    lambda = 1, 
    from = 0, 
    to = 10,
    by = 0.01
  ) {
  checkmate::assert_number(s, lower = 0)
  checkmate::assert_number(i, lower = 0)
  checkmate::assert_number(r, lower = 0)
  checkmate::assert_number(beta)
  checkmate::assert_number(lambda)
  checkmate::assert_number(from, lower = 0)
  checkmate::assert_number(to, lower = from)
  checkmate::assert_number(by, lower = 0)
  
  fun <- function (t, y, parms) {
    list2env(as.list(y), envir = environment())
    list2env(as.list(parms), envir = environment())
    
    list(
      c(
        ds = (- beta) * s * i,
        di = beta * s * i - lambda * i,
        dr = lambda * i
      )
    )
  }
  
  initial_values <- c(s = s, i = i, r = r)
  parameters <- list(beta = beta, lambda = lambda)
  time <- seq(from = from, to = to, by = by)
  
  data <- 
    deSolve::ode(
      y = initial_values,
      times = time, 
      func = fun,
      parms = parameters
    ) |>
    dplyr::as_tibble() |>
    dplyr::mutate(dplyr::across(dplyr::everything(), ~ as.numeric(.x)))
  
  list(
    data = data,
    initial_values = as.list(initial_values),
    parameters = as.list(parameters)
  ) |>
  invisible()
}
```

```{r}
sir() |> magrittr::extract2("data")
```

## Plotting disease dynamics

```{r}
plot_pop_dynamics <- function(
    s = 1, 
    i = 0.05, 
    r = 0, 
    beta  = 3, 
    lambda = 1, 
    from = 0, 
    to = 10,
    by = 0.01
  ) {
  checkmate::assert_number(s, lower = 0)
  checkmate::assert_number(i, lower = 0)
  checkmate::assert_number(r, lower = 0)
  checkmate::assert_number(beta)
  checkmate::assert_number(lambda)
  checkmate::assert_number(from, lower = 0)
  checkmate::assert_number(to, lower = from)
  checkmate::assert_number(by, lower = 0)
  
  sir(s, i, r, beta, lambda, from, to, by) |> list2env(envir = environment())
  
  plot <- 
    data |>
    ggplot2::ggplot(ggplot2::aes(x = time)) +
    ggplot2::geom_line(
      ggplot2::aes(y = s, color = "Susceptible"),
      linewidth = 0.75
    ) +
    ggplot2::geom_line(
      ggplot2::aes(y = i, color = "Infected"),
      linewidth = 0.75
    ) +
    ggplot2::geom_line(
      ggplot2::aes(y = r, color = "Recovered"),
      linewidth = 0.75
    ) +
    ggplot2::labs(
      title = "SIR Model Disease Dynamics",
      subtitle = latex2exp::TeX(
        paste0(
          "$S_0$ = ", s, " | ",
          "$I_0$ = ", i, " | ",
          "$R_0$ = ", r, " | ",
          "$\\beta$ = ", round(beta, 2), " | ",
          "$\\lambda$ = ", round(lambda, 2)
        ),
      ),
      x = "Time", 
      y = "Proportion",
      color = ggplot2::element_blank()
    ) +
    ggplot2::scale_color_manual(
      breaks = c("Susceptible", "Infected", "Recovered"),
      values = c("blue", "red", "black")
    )
  
  print(plot)
  
  invisible()
}
```

```{r}
plot_pop_dynamics()
```

## Phase space visualization

```{r}
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
```

```{r}
plot_phase_space <- function(
    s = 1, 
    i = 0.05, 
    r = 0, 
    beta = seq(3, 8, by = 1), 
    lambda = 1, 
    from = 0, 
    to = 100,
    by = 0.01,
    theta = 180,
    phi = 0
  ) {
  checkmate::assert_number(s, lower = 0)
  checkmate::assert_number(i, lower = 0)
  checkmate::assert_number(r, lower = 0)
  checkmate::assert_numeric(beta)
  checkmate::assert_number(lambda)
  checkmate::assert_number(from, lower = 0)
  checkmate::assert_number(to, lower = from)
  checkmate::assert_number(by, lower = 0)
  checkmate::assert_number(theta, lower = 0)
  checkmate::assert_number(phi, lower = 0)

  plot <-ggplot2::ggplot()
  
  for (j in beta) {
    data_j <- 
      sir(s, i, r, j, lambda, from, to, by) |>
      magrittr::extract2("data") |>
      dplyr::mutate(color = as.character(j))
    
    plot <-
      plot +
      ggplot2::geom_path(
        data = data_j,
        ggplot2::aes(x = r, y = s, color = color),
        linewidth = 0.75
      )
  }
  
  colors <- gg_color_hue(length(beta))
  names(colors) <- as.character(beta)
  
  plot <-
    plot +
    ggplot2::labs(
      title = "SIR Model Phase-Space",
      subtitle = latex2exp::TeX(
        paste0(
          "$S_0$ = ", s, " | ",
          "$I_0$ = ", i, " | ",
          "$R_0$ = ", r, " | ",
          "$\\lambda$ = ", round(lambda, 2)
        ),
      ),
      x = "Recovered",
      y = "Susceptible"
    ) +
    scale_color_manual(
      name = latex2exp::TeX("$\\beta$"), 
      values = colors
    )
  
  print(plot)
  
  invisible()
}
```

```{r}
plot_phase_space()
```

## References {.unnumbered}

::: {#refs}
:::
